A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-09 tidyindex.qmd
changes after morning re-read

The tidy framework consists of two key components: tidy data and tidy tools. The concept of tidy data [@wickham_tidy_2014] prescribes specific rules for organising data in an analysis, with observations as rows, variables as columns, and types of observational units as tables. Tidy tools, on the other hand, are concatenated in a sequence through which the tidy object flows, creating a pipeline for data processing. These pipelines are data centric, meaning all the functions take a tidy data object as inputs and return a processed tidy data object, directly ready for the next operations without further reshaping needed. Also, the pipeline approach corresponds to the modular programming practice, which break down complex problems into smaller and more manageable pieces, as oppose to a monolithic design, where all the steps are predetermined and integrated into a single piece.

@@

Examples of using a pipeline approach to structure data analysis in R can be traced back to the interactive graphics literature in the earlier time [@buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014; @wickham_plumbing_2009]. @wickham_plumbing_2009 argues that whether made explicit or not, pipeline has to be presented in every graphics program and making them explicit is beneficial for understanding the implementation and comparing between different graphic systems. While this comment is made in the context of graphics program, it is also applicable to other general or domain specific data analysis workflow. More recently, the tidyverse suite [@wickham_welcome_2019] takes the pipeline approach for general-purpose data wrangling and has gained popularity within the R community. The pipeline-style code can be directly read as a series of operations applied successively on the tidy data object, offering a method to document the data wrangling process. Unlike textual documentation, which may omit seemingly obvious or less significant information, the pipeline approach captures all the computational details while retaining the conciseness of a coding language, making it easy for human analysts to comprehend.

Since the success of tidyverse, more packages have been developed to analyze data using the tidy framework for domains specific applications, a noticeable example of which is `tidymodels` for building machine learning models [@tidymodels]. To create a tidy workflow tailored to a specific domain, developers first need to identify the fundamental building blocks in the typical workflow. These components are then implemented as modules, which can be combined to form the pipeline. For example, in supervised machine learning models, steps such as data splitting, model training and model evaluation are commonly used in most workflow. In the `tidymodels`, these steps are correspondingly implemented as package `rsample`, `parsnip`, and `yardstick`, agnostic to the specific model chosen.

@@

For constructing indexes, rather than implementing each index separately, as has been done for some indexes, the pipeline approach introduces explicit, user-facing modules that analysts can compose into indexes. This approach provides several advantages to analysts and researchers working on indexes. The index construction process can be complicated to describe and readable pipeline code complements conventional diagrams and textual description for understand the computational details. The uniform interface allows analysts to assemble a diverse array of indexes with a shared set of components, reducing the learning curve for adopting new implementation for different indexes. Importantly, each pipeline module and its corresponding steps are implemented as standalone functions, providing great flexibility for analysis the index. Analysts can rearrange the order in the pipeline or insert new step to experiment with possible new definition for the index or to understand index behaviour.

>>>>

The tidy framework consists of two key components: tidy data and tidy tools. The concept of tidy data [@wickham_tidy_2014] prescribes specific rules for organising data in an analysis, with observations as rows, variables as columns, and types of observational units as tables. Tidy tools, on the other hand, are concatenated in a sequence through which the tidy object flows, creating a pipeline for data processing. These pipelines are data centric, meaning all the functions take a tidy data object as inputs and return a processed tidy data object, directly ready for the next operations to be applied. Also, the pipeline approach corresponds to the modular programming practice, which breaks down complex problems into smaller and more manageable pieces, as oppose to a monolithic design, where all the steps are predetermined and integrated into a single piece. The flexibility provided by the modularity makes it easier to modify certain steps in the pipeline and to maintain and extend the code base.

@@

Examples of using a pipeline approach to structure data analysis in R can be traced back to the interactive graphics literature in the earlier time [@buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014; @wickham_plumbing_2009]. @wickham_plumbing_2009 argues that whether made explicit or not, pipeline has to be presented in every graphics program and making them explicit is beneficial for understanding the implementation and comparing between different graphic systems. While this comment is made in the context of graphics program, it is also applicable generally to data analysis workflow. More recently, the tidyverse suite [@wickham_welcome_2019] takes the pipeline approach for general-purpose data wrangling and has gained popularity within the R community. The pipeline-style code can be directly read as a series of operations applied successively on the tidy data object, offering a method to document the data wrangling process for analysts to comprehend. Unlike textual documentation, which may omit seemingly obvious or less significant information, the pipeline approach captures all the computational details while retaining conciseness.

Since the success of tidyverse, more packages have been developed to analyze data using the tidy framework for domains specific applications, a noticeable example of which is `tidymodels` for building machine learning models [@tidymodels]. To create a tidy workflow tailored to a specific domain, developers first need to identify the fundamental building blocks in the typical workflow. These components are then implemented as modules, which can be combined to form the pipeline. For example, in supervised machine learning models, steps such as data splitting, model training and model evaluation are commonly used in most workflow. In the `tidymodels`, these steps are correspondingly implemented as package `rsample`, `parsnip`, and `yardstick`, agnostic to the specific model chosen. The uniform interface in tidymodels frees analysts from recalling model-specific syntax for performing the same operation across different models, increasing the efficiency to work with different models simultaneously.

@@

For constructing indexes, rather than implementing each index separately, as has been done for some indexes, the pipeline approach introduces explicit, user-facing, and standalone modules for analysts to assemble indexes. Analysts can rearrange the order in the pipeline or insert new step to experiment with possible new index definition or to understand index behaviour. Also, the readability of the pipeline code can complement conventional diagrams and textual information to understand the how indexes are constructed. The uniform interface of pipeline allows analysts to assemble a diverse array of indexes with a shared set of components, reducing the learning curve for adopting new implementation for different indexes.

### 2023-08-09 tidyindex.tex
changes after morning re-read

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden, enhanced]}{\end{tcolorbox}}\fi

@@

The tidy framework consists of two key components: tidy data and tidy
tools. The concept of tidy data (Wickham 2014) prescribes specific rules
for organising data in an analysis, with observations as rows, variables
as columns, and types of observational units as tables. Tidy tools, on
the other hand, are concatenated in a sequence through which the tidy
object flows, creating a pipeline for data processing. These pipelines
are data centric, meaning all the functions take a tidy data object as
inputs and return a processed tidy data object, directly ready for the
next operations without further reshaping needed. Also, the pipeline
approach corresponds to the modular programming practice, which break
down complex problems into smaller and more manageable pieces, as oppose
to a monolithic design, where all the steps are predetermined and
integrated into a single piece.

Examples of using a pipeline approach to structure data analysis in R
can be traced back to the interactive graphics literature in the earlier
time (Buja et al. 1988; Sutherland et al. 2000; Xie, Hofmann, and Cheng
2014; Wickham et al. 2009). Wickham et al. (2009) argues that whether
made explicit or not, pipeline has to be presented in every graphics
program and making them explicit is beneficial for understanding the
implementation and comparing between different graphic systems. While
this comment is made in the context of graphics program, it is also
applicable to other general or domain specific data analysis workflow.
More recently, the tidyverse suite (Wickham et al. 2019) takes the
pipeline approach for general-purpose data wrangling and has gained
popularity within the R community. The pipeline-style code can be
directly read as a series of operations applied successively on the tidy
data object, offering a method to document the data wrangling process.
Unlike textual documentation, which may omit seemingly obvious or less
significant information, the pipeline approach captures all the
computational details while retaining the conciseness of a coding
language, making it easy for human analysts to comprehend.

Since the success of tidyverse, more packages have been developed to
analyze data using the tidy framework for domains specific applications,
a noticeable example of which is \texttt{tidymodels} for building
machine learning models (Kuhn and Wickham 2020). To create a tidy
workflow tailored to a specific domain, developers first need to
identify the fundamental building blocks in the typical workflow. These
components are then implemented as modules, which can be combined to
form the pipeline. For example, in supervised machine learning models,
steps such as data splitting, model training and model evaluation are
commonly used in most workflow. In the \texttt{tidymodels}, these steps
are correspondingly implemented as package \texttt{rsample},
\texttt{parsnip}, and \texttt{yardstick}, agnostic to the specific model
chosen.

For constructing indexes, rather than implementing each index
separately, as has been done for some indexes, the pipeline approach
introduces explicit, user-facing modules that analysts can compose into
indexes. This approach provides several advantages to analysts and
researchers working on indexes. The index construction process can be
complicated to describe and readable pipeline code complements
conventional diagrams and textual description for understand the
computational details. The uniform interface allows analysts to assemble
a diverse array of indexes with a shared set of components, reducing the
learning curve for adopting new implementation for different indexes.
Importantly, each pipeline module and its corresponding steps are
implemented as standalone functions, providing great flexibility for
anlaysis the index. Analysts can rearrange the order in the pipeline or
insert new step to experiment with possible new definition for the index
or to understand index behaviour.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, frame hidden, boxrule=0pt, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

The tidy framework consists of two key components: tidy data and tidy
tools. The concept of tidy data (Wickham 2014) prescribes specific rules
for organising data in an analysis, with observations as rows, variables
as columns, and types of observational units as tables. Tidy tools, on
the other hand, are concatenated in a sequence through which the tidy
object flows, creating a pipeline for data processing. These pipelines
are data centric, meaning all the functions take a tidy data object as
inputs and return a processed tidy data object, directly ready for the
next operations to be applied. Also, the pipeline approach corresponds
to the modular programming practice, which breaks down complex problems
into smaller and more manageable pieces, as oppose to a monolithic
design, where all the steps are predetermined and integrated into a
single piece. The flexibility provided by the modularity makes it easier
to modify certain steps in the pipeline and to maintain and extend the
code base.

Examples of using a pipeline approach to structure data analysis in R
can be traced back to the interactive graphics literature in the earlier
time (Buja et al. 1988; Sutherland et al. 2000; Xie, Hofmann, and Cheng
2014; Wickham et al. 2009). Wickham et al. (2009) argues that whether
made explicit or not, pipeline has to be presented in every graphics
program and making them explicit is beneficial for understanding the
implementation and comparing between different graphic systems. While
this comment is made in the context of graphics program, it is also
applicable generally to data analysis workflow. More recently, the
tidyverse suite (Wickham et al. 2019) takes the pipeline approach for
general-purpose data wrangling and has gained popularity within the R
community. The pipeline-style code can be directly read as a series of
operations applied successively on the tidy data object, offering a
method to document the data wrangling process for analysts to
comprehend. Unlike textual documentation, which may omit seemingly
obvious or less significant information, the pipeline approach captures
all the computational details while retaining conciseness.

Since the success of tidyverse, more packages have been developed to
analyze data using the tidy framework for domains specific applications,
a noticeable example of which is \texttt{tidymodels} for building
machine learning models (Kuhn and Wickham 2020). To create a tidy
workflow tailored to a specific domain, developers first need to
identify the fundamental building blocks in the typical workflow. These
components are then implemented as modules, which can be combined to
form the pipeline. For example, in supervised machine learning models,
steps such as data splitting, model training and model evaluation are
commonly used in most workflow. In the \texttt{tidymodels}, these steps
are correspondingly implemented as package \texttt{rsample},
\texttt{parsnip}, and \texttt{yardstick}, agnostic to the specific model
chosen. The uniform interface in tidymodels frees analysts from
recalling model-specific syntax for performing the same operation across
different models, increasing the efficiency to work with different
models simultaneously.

For constructing indexes, rather than implementing each index
separately, as has been done for some indexes, the pipeline approach
introduces explicit, user-facing, and standalone modules for analysts to
assemble indexes. Analysts can rearrange the order in the pipeline or
insert new step to experiment with possible new index definition or to
understand index behaviour. Also, the readability of the pipeline code
can complement conventional diagrams and textual information to
understand the how indexes are constructed. The uniform interface of
pipeline allows analysts to assemble a diverse array of indexes with a
shared set of components, reducing the learning curve for adopting new
implementation for different indexes.
